Convergent Evolution During Local Adaptation to Patchy Landscapes

### 2014-11-26 patchy-review-responses.tex

% \documentclass{article}
% \usepackage{fullpage}
% \input{review-response-commands}
% 
% \begin{document}

\section*{From the editor}
% From: "Catriona J. MacCallum" <cmaccallum@plos.org>
% To: Peter Ralph <pralph@usc.edu>
% Date: Fri, 15 Feb 2013 04:01:05 -0500
% Subject: PLOS Biology Decision [PBIOLOGY-D-12-02842R1]
% Dear Dr. Ralph,

\begin{minipage}[b]{2.5in}
  Resubmission Cover Letter \\
  {\it PLoS Genetics}
\end{minipage}
\hfill
\begin{minipage}[b]{2.5in}
    Department of Molecular and Computational Biology\\
    University of Southern California, \\
    \emph{and} Department of Evolution and Ecology\\
    University of California at Davis\\
  \today
\end{minipage}

@@

%%%%%%
\point{}{
    Am I right that the basic argument is the
    following:

    A migrant family in unfavorable territory will be typically of size $n \sim 1/s$
    ($s=s_m$), stick around for a time $\tau \sim 1/s$ and be spread out over $\ell \sim \sigma/\sqrt(s)$. 
    Given an equilibrium density $\rho*q(x)$, we need an arrival rate
    of migrant families of
    \[ k ~ \rho q(x) l/n/\tau \sim \rho q(x) l s^2 \]
}

\reply
The factor of $\ell$ shouldn't be there:
suppose that $X$ is a Poisson process of ``centers'' of rate 1, 
and that around each center of $X$ we distribute $n$ points uniformly within a circle of radius $\ell$;
the expected number of points in a set of area 1 is $n$, regardless of $\ell$.
The right thing to do with the factor of $\tau$ is harder to intuit:
it is clear that $\rho q(x)$ should be $E[K]$ multiplied by the number of families present;
and it might be argued that the number of families present should be the influx of families
multiplied by the time they stick around for,
but given that the typical size of a family decreases once it arrives,
the correct way forward becomes non-intuitive, to us, at least.

%%%%%%%%
\point{}{ 
    The probability of establishment of at least one member of family is then
    \[ P_{est} = 1 - \exp(-p n) ~ 1 - \exp(-p_e/s) \]
    where $p_e$ is the establishment probability of a single individual. The result
    for $\migrate$ will then depend on whether $p_e/s$ is large or small.
    For $p_e/s$ large, $\migrate = \rho q(x) s^2 \ell$,
    otherwise,
    $\migrate = \rho q(x) s \ell p_e$.

@@

    The latter is the case discussed in the manuscript. The other is important too
    and should be discussed. 
}

%%%%%%%%
\point{}{ 
    I don't seem to be able to identify the equivalent of
    $\ell$ in the author's formulae, but I believe it should be there. After all, 
    $\rho q(x)$ is a density and needs to be multiplied by a length (or area in d=2) to
    give a meaningful rate.
}

@@

%%%%%%%%
\point{Fig 4:}{ Why not use natural units for the axis of these plots,
  $R\sqrt(s)/\sigma$, $\rho \mu A/s$.
  }

%%%%%%%%
\point{in the discussion on page 10 "probability of parallel adaptation.."}{
    it should be noted that many of the scenarios at high mutation influx are quite
    hypothetical. While the colonization of migrants might dominate over local
    adaptation, this argument assumes that the migrants are at equilibrium
    frequency. The transient to reach this equilibrium might be much longer than
    the time required for de novo local adaptation.
}

%%%%%%%%%%%%%%  Reviewer 2
\reviewersection

\begin{quote}
Briefly, I think that the derivation of the main results (12), (21), and (25) can
be made much shorter and easier to understand (for both empiricists and the-
orists), and that there need to be simulations that quantitatively confirm these
results. Overly-detailed comments are below.
\end{quote}

\point{}{
    I think that the main result (12) can be derived much more simply, and
    also extended. By writing that the probability that a family that arrives
    at the patch is successful is approximately $ 2s_p \E [K] /\xi^2$, the authors have
    implicitly assumed that this quantity is $<\ll 1$. But this implies that the
    fluctuations in the number of B individuals arriving at the patch can be
    ignored, since even a large fluctuation does not saturate the probability
    of success. This is valid for $s_m \gg s_p$. Given that the fluctuations can be
    ignored, (12) can be derived immediately from $q(y)$ -- simply apply Fick’s
    Law to $q(y)\rho$ to get that the flux of new B individuals is $\sigma^2 |q'(y)|\rho/2$, and
    then multiply by the success probability $2s_p/\xi^2$. (Actually, the result is
    slightly different from (12) -- see X below.)
}

@@

\reply
This argument falls in the trap of treating each migrant individual as independent,
ignoring the correlations of arrival times induced by families;
see the discussion in which we now set this up as a straw man in XXX.

\point{}{
    However, I think we should also care about the opposite limit, $s_m \ll s p$. In
    this case, the fluctuations do matter -- arrivals typically occur in clumps
    (families) large enough so that success is almost certain conditioned on
    arrival. Since on average each family consists of $\E[K]$ individuals, the
    family flux is just the individual flux divided by $\E[K]$. $\E[K] = \xi^2 /(2s_m)$,
    so the result is the same as above, except with $s_p$ replaced by an additional
    factor of $s_m$. (This could be made more formal, but I don’t think that’s
    what the authors are going for.) In the intermediate case $s_m \approx s_p$ which
    might be the most relevant for the mice, the two limiting regimes cross
    over smoothly. So for the first limit ($s_m \gg s_p$), all you need is $q$, and
    for the second, you only need to add basic facts about branching random
    walks (ie, limiting number distribution and localization at long times).
}

@@

\reply
In this argument, the reviewer is taking into account clumpiness for establishment of migrants,
but still not for the time interval
between arrivals of families.

%%%%%%%%
\point{}{ 
    As it stands, $\migrate(y)$ is the rate of successful migrants at an infinitesimal
    area at distance $y$, so it needs to be integrated over an area to get the rate
    for a whole patch. The authors don’t do this, and as a result, expressions
    like (14) are dimensionally inconsistent. It’s a little annoying to define the
    right area of integration, but it’s straightforward to argue that up to a
    constant the answer
     should be given by integrating $\migrate(y)$ over a region
     of thickness $\sigma/\sqrt{2sm}$ around the boundary of the patch. I think it’s easier
    to work with fluxes and just integrate over the boundary of the patch, with
    no worries about thickness, so that’s why I used them above. (Of course,
    you get the same answer either way, except for small or weirdly-shaped
    patches, where either or both can break down.)
}

%%%%%%%%
\point{}{ 
    I think that the authors need to show simulation results confirming the
    analytic approximations. Spatial simulations are a pain, but it’s important
    to have a (somewhat) independent check on the math and assumptions. At
    a minimum, I think that there needs to be a figure showing that they have
    the right expression for the rate of successful migrants (12) and migration
    sojourn times (17) over a range of parameter values. If it’s not too much
    work, I think it would be really nice to have a figure that doubles as a kind
    of graphical summary, showing a scattering of patches in two dimensions,
    with one starting as the homeland of the B allele and the rest colored
    according to their probability in simulations of being colonized by that
    original allele vs new mutations, and a curve marking the region predicted
    as being colonized by the analytical approximations, or something like
    that. But I understand if the authors don’t want to spend their time on
    that.
}

%%%%%%%%
\point{}{ 
    It would be nice to have a table of symbol definitions.
}

\reply
We've tried to say in words what each symbol is when we use it as much as possible,
so we hope this isn't necessary.

%%%%%%%%
\point{}{ 
    And speaking of symbols, I think it’s a bit nicer to write things in terms
    of the characteristic patch length $a \sim A^{1/d}$, rather than the area. This
    avoids having to talk about one-dimensional areas \ldots
}

\reply
Good point, but $A$ is what shows up naturally in $\mutrate$, so we've left it as-is.

%%%%%%%%
\point{}{ 
     It would be nice to have a clear, concise summary of the quantitative
     assumptions needed for the results (all the stuff like $a \gg \sigma/\sqrt{2s_m}$ or
     whatever).
 }

\reply
See text added in \nameref{ss:assumptions}.

%%%%%%%%
\point{"Length of the hitchhiking haplotype"}{
    I would focus more on the asymptotic expressions for $R \gg w  \gg \sigma/ \sqrt{2s_m}$, since that’s the regime
     where the analysis makes the most sense. Then the expressions become
     simpler and you can treat d = 2 as well. If you want, you could even
     lead with the intuitive explanation at the end of the section, and save the
     Bessel functions for an appendix.
}

%%%%%%%%
\point{}{ 
     I think (24) follows directly from Barton, Heredity 1979. (If you look at
     the unnumbered equation above Fig. 1 in that paper and plug in $x \sim a$,
     you find the time scale $r_{eff}$.) I think that the displayed equation should just
     be $r_{eff} \approx r\sigma/\sqrt{s}/a$. (I feel like the extra correction terms and geometric
      factor go against the spirit of the paper, and in any case are probably
      small compared to the inaccuracies introduced just by applying the model
      to any real system.) I think you can cut essentially everything before
      the simple explanation just above (24); if people want the mathematical
      details, they can read Barton. If I am misunderstanding things and there
    is a new result here, rather than just a good new exposition of an existing
    result, there should be a figure showing simulations confirming it.
}

%%%%%%%%
\point{In “Applications”,}{the authors spend a lot of time introducing the mice
    and calculating the values for Fig. 6, but then they don’t show where the
    actual data lie on those plots. Is there enough information to do some
    kind of comparison? How close are the nearest samples to each other?
    Have no examples of alleles spreading by migration been found? If not,
    how much of a bound does this put on the parameter values? Is there any
    data on the genetic map length of the region of reduced diversity around
    the selected locus?
}

%%%%%%%%
\point{}{ 
    A few sections or passages seem out of place:
    \begin{itemize}
        \item[A.]  I think “Simulation methods” should go in the “Materials and meth-
            ods” section, or at least not right between “Establishment of a locally
            adaptive allele due to mutational influx” and “Establishment of a lo-
            cally adaptive allele due to mutational influx”.

        \item[B.] The sentence on p7 beginning “By our definition of fitness. . . ”.

        \item[C.]  The paragraph starting “Also note that. . . ” on p11.
    \end{itemize}
}

%%%%%%%%
\point{Fig. 4:}{$R/\sigma$ should have dimensions of $\sqrt{\text{time}}$. In the caption, I think it’s
    misleading to say that $\migrate$ is ``relatively insensitive to $s_m$''. It’s exponen-
    tially sensitive! I don’t understand the blue numbers in the right-hand
    panel. It might be easier to interpret these if they were contour plots
    of the total time to adaptation, colored according to the value of (14),
    although I’m not sure. Also, they aren’t really ``phase diagrams''.
}

%%%%%%%%
\point{Top of p11 (just above “Multiple patches”):}{should be $\sqrt{2s_m}/\sigma$.
}

%%%%%%%%
\point{Between (49) and (51):}{should be “Green’s”.
}

%%%%%%%%
\point{(51)-(55)}{are missing factors of $\sigma$.
}

%%%%%%%%
\point{}{ 
 In (56) and following, it would be clearer to use different symbols for scaled
and unscaled distances.

}

@@

}

\reply

%%%%%%%%%%%%%%  Reviewer 3
\reviewersection

\point{1a.}{
        What I miss is a clearer discussion of the scope and limits of the
    main results. Take Eq. 13: in its revised form, this is a non-monotonic
    function of the selection coefficient $s_m$. I am almost sure that this is
    an artefact and that the parameter range of the theory needs to be
    restricted to the part where the transition probability declines with $s_m$.
    But it is not easy to see how this needs to be done in a proper way. We
    certainly need a minimal distance among patches. On top of that, we may
    need a lower bound for $s_m$. Both these restrictions (or any other) should
    not be just ad-hoc, but motivated from within the theory (see my remarks at eq. 13).
}

@@

\point{3.}{
    Section "length of shared haplotype": This is very handwavy and
    (without simulation backup) I find it hard to believe that these results
    are accurate. Since it is also not central to the paper, the section
    should maybe just be dropped.
}

%%%%%%
\point{[beginning comments in the pdf]}{
``Do you assume a Gaussian kernel? Or is this irrelevant? Which properties of the kernel are required? Only existence of a finite $\sigma^2$?''
}

\reply
We've tried to make it more clear here 
\llname{rr:longtailed}
exactly what we sweep under the rug later.

%%%%%%
\point{}{``This is not sufficient: the patch also needs to be large enough that $p_e$ is (almost) reached at its center. Generally, the approximation should over-estimate the true value.''}

@@

\reply
Fixed? \llname{rr:awkward}

%%%%%%%%
\point{[expected total number]}{if y is a point, this number would be 0 (its a density)}

\reply
We have added the word ``informally''. \llname{rr:density}

%%%%%%%%
\point{[equation \eqref{eqn:migrate}]}{ 
    \begin{quote}
    This function is non-monotonic in $s_m$ and I think this needs to be discussed: can it really be that $\migrate$ increases with $s_m$ or is this an artifact?

    In general, we need a clearer idea about the limits of this theory. There are several things to consider:
    \begin{itemize}
        \item equation (3) for the expected mutant density is only asymptotic. Therefore, we probably need to require $|y| \gg \sigma / \sqrt(2s_m)$.

@@

        \item over short ranges, results must depend on details of the dispersal kernel anyway. [Depending on the shape of the distribution - heavy tail - this even holds for large distances]. As mentioned above: spell out assumptions on the kernel.
    \end{itemize}

    So: what is the minimal distance $|y|$ for the result to be valid? And: does this minimal distance eliminate the anomaly of an increasing $\migrate$ with $s_m$ for small $s_m?$ Alternatively: is there also a lower limit for $s_m$? I have noted that you do assume above that $s_m$ is large enough to avoid local fixation - but I'm not sure that this comes down to the same thing. 
\end{quote}
}

\reply

%%%%%%%%
\point{[expression for the outflux of migrants]}{do we need this for anything?}

@@

\reply
We don't need it for anything else, 
but we think it adds sufficiently to the intuition about the process
that we'd like to keep it in.
In practice it could provide a good reality check.
\plr{alternatively, compute this for the mice example?}

%%%%%%%%
\point{[figure 1 showing unimportance of square root term]}{}

\reply
Clarified. \llname{rr:sqrt_term}

%%%%%%%%
\point{``the assumption of well-separated patches''}{ 
}

\reply
We have changed this. \llname{rr:separated_patches}

%%%%%%%%
\point{[figure 4, the contour plots]}{
}

\reply
Good point; we didn't notice that after correcting $\migrate$ the parameter regime on the right didn't make sense.
We have substituted XXX for these.

%%%%%%%%
\point{[\eqref{eqn:total_migrate} assumes no interference}{ 
}

\reply
Good point, but one we don't think we need to get into; we've added a small caveat.
\llname{rr:interference}

%%%%%%%%
\point{}{ 
    One message here seems to be that "migration vs mutation" is independent of the density. As far as the model goes, this is correct. However, in Nature selection will typically be density dependent, usually with stronger deleterious selection (higher $s_m$) at higher densities. One could therefore suspect that higher densities will often favor mutation relative to migration.
}

\reply
Interesting point,
but we think its applicability depends on what densities are being compared.
If it is in two different species, then $s_m$ must clearly be measured separately in the two,
so this is not a worry.
If it is two populations of the same species,
then one might also argue that lower densities implies worse habitat,
and so stronger deleterious selection.
We don't see anything straightforward we can say here, 
other than adding this caveat later. \llname{rr:popdensity}

@@

%%%%%% p.12
\point{[justify $\E[K]\approx 1/s_m$]}{}

\reply
See \eqref{eqn:mean_K}, and modified discussion around \llname{rr:EK}.

%%%%%%%%
\point{}{ 
    There is a lot of handwaving in this section. Although each single step has some plausability, the final result seems to depend on a lot of assumptions. There is also an important difference to the other sections. Most of the other results are valid for large distances, where we can safely use asymptotic forms of the distribution (which arguably are quite universal). In contrast, this section relies on short-range dispersal (limited excursions of B-mutant lines to the surrounding area, where b is common). This is far less universal and should depend on details of dispersal.

    Also, it seems to me that another route to break down LD is by b-immigrants to the patch, who will life there for an exponential life span and can recombine before they die out. Note that, in the pocket mouse example below, the frequency of dark mice on the lava flows is only 86\%. This should result in considerable recombination "on site".

    So, in summary: I am sceptical concerning the results of this section. They are preliminary at best and need further study. I therefore suggest to cut them from the paper (or reduce it to some brief remarks).
}

%%%%%%%%
\point{Dominance.}{ 
    the deleterious effect could be recessive and the beneficial one still (at least partially) domninant.
}

\reply
Added that note. \llname{rr:dominance}

### 2014-11-26 patchy-selection-paper.tex
more responses...

%%%%%
% Responses to reviews:
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
% \reviewresponsestrue  % include them
\ifreviewresponses\else\renewcommand{\linelabel}[1]{}\fi
\reviewresponsesfalse  % don't include them
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

@@

% \newcommand{\var}{\mathop{\mbox{Var}}}
% \newcommand{\Exp}{\mathop{\mbox{Exp}}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}

@@

Let $p(x)$ be the probability that a new mutant $B$ allele that arises at location $x$
relative to the center of the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is close to $p_e$ at some point and is symmetric about the edge of the patch. \lineref{rr:p_estab}
We examine this more generally via exact calculation of $p(x)$ in the section \nameref{apx:establishment_sims}.

@@

Since each individual has on average $e^{-s_m}$ offspring,
the expected family size after time $t$ is $e^{- t s_m}$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $e^{-s_m t} = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed 
(see \citet{jagers1975branching} for details),
then $K_t$ has a limiting distribution: $K_t \dconv K$,
and 
\begin{align} \label{eqn:EK_limit}
    e^{-s_m t}/(1-k_e(t)) \to \E[K]$ as $t \to \infty .
\end{align}
In other words, the chance of survival to time $t$ is a constant multiplied by $e^{-s_m t}$,
and conditional on survival, the family size is given by a fixed distribution.

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and using that $\E[K] \approx 1/s_m$, 
\begin{align} \label{eqn:twocase_migrate}
    \migrate(x) &\approx 2 \rho s_m \min(s_m,p_e) \times q(x) .
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this, we set $p_e = 2 s_p / \xi^2$.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation) 
it becomes established in the poor habitat patch rapidly
if it escapes loss by demographic stochasticity.
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
To work out how adaptation is likely to proceed, 
we can compare expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above
for the effective rates of adaptation by new mutation and by migration.
We work in one dimension, as the square root term appearing for two dimensions is relatively unimportant (see figure~\ref{fig:sim_occupation_freqs}).

@@

If the mutational influx is higher -- say, $\rho A \mu = .01$ --
the required separation between patches is only reduced to $R/\sigma > 7.3$ with $s_m=.05$.
If $s_m=.001$, then with this higher mutational influx this calculation predicts that mutation will \emph{always} be faster than migration
-- however, this is misleading, 
since with $s_m=.001$, for the assumption of well-separated patches to hold we would need $R/\sigma > 1/\sqrt{2s_m} = 22.4$.
The correct conclusion is that convergent adaptation is much more likely than shared adaptation
between patches separated by at least 22.4$\sigma$;
for shorter separations the model breaks down.

@@

\begin{figure}[ht]
  \begin{center}
    \includegraphics{phase-diagram-log}
  \end{center}
  \caption{
  Shaded contour plots of
  mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
  as a function of population density ($\rho$), 
  distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
  strength of deleterious selection ($s_m$),
  and mutation rate scaled by patch area ($\mu A$).
  The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
  The black, dotted line shows values where $\Tmig=\Tmut$.
  Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
  Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
  \label{fig:phase_diagram}
  }
\end{figure}

@@

If not yet adapted, a focal patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $R_1, \ldots, R_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}, is
\begin{equation} \label{eqn:total_migrate}
  \sum_{i=1}^k \migrate(R_i) = C \frac{ 4 \rho s_p s_m }{\xi^2} \sum_{i=1}^{k} \left(R_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m} /\sigma\right).
\end{equation}
If the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}   \label{eqn:migprob}
  \frac{\left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
Equations \eqref{eqn:total_migrate} and \eqref{eqn:migprob} specify the transition rates of the process.
Since the compound parameter $s_p \rho / \xi^2$ is common to both migration and mutation rates,
it functions only as a time scaling of the process, 
and therefore has no effect on the final configuration, namely, 
the sets of patches sharing the genetic basis of their adaptive response.

@@

>>>>

%%%%%
% Responses to reviews:
\input{review-response-commands}
% set this to show line numbers and include responses to reviews or not
\newif\ifreviewresponses
\reviewresponsestrue  % include them
% \reviewresponsesfalse  % don't include them
\ifreviewresponses\else\renewcommand{\linelabel}[1]{}\fi
\iflatexml\reviewresponsesfalse\fi  % they don't work so well in html
% \newcommand{\responsefile}{pbio-reviews-19sept12-responses.tex}  % name of the review reponses file
\newcommand{\responsefile}{patchy-review-responses.tex}  % name of the review reponses file

@@

% \newcommand{\var}{\mathop{\mbox{Var}}}
% \newcommand{\Exp}{\mathop{\mbox{Exp}}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\Exp}{Exp}
\DeclareMathOperator{\sn}{sn}
\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\one}{\mathbf{1}}
\newcommand{\calE}{\mathcal{E}}
\newcommand{\calC}{\mathcal{C}}
\newcommand{\dconv}{\xrightarrow{d}}
\newcommand{\deq}{\stackrel{\scriptscriptstyle{d}}{=}}
\newcommand{\given}{\,\vert\,}

@@

Let $p(x)$ be the probability that a new mutant $B$ allele that arises at location $x$
relative to the center of the patch fixes within the poor habitat patch.
Under various assumptions, precise expressions can be found for $p(x)$ \citep{barton1987establishment},
% the function $p(x)$ can be found (in one dimension) in terms of Jacobi elliptic functions,
but results will be more interpretable if we proceed with a simple approximation.
The total influx per generation of mutations that successfully establish is given by the
integral of $p(x)$ over the entire species range:
\begin{equation}
  \mutrate = \rho \mu \int p(x) dx \label{eqn:exactmutrate}
\end{equation}
% $\int \rho \mu p(x) dx$,
and hence depends in a complicated way on the patch geometry and selection coefficients,
but still scales linearly with the mutational influx density $\rho \mu$.
If we consider a patch of area $A$, whose width is large relative to $\sigma/\sqrt{2s_m}$, 
then a reasonable approximation is to ignore migration, 
so that $p(x) = p_e \approx 2 s_p / \xi^2$ within the patch, and $p(x) = 0$ otherwise.
%for a patch of area $A$ approximating $\int p(x) dx \approx 2 s_p A / \xi^2$. 
This approximates the integrand $p(x)$ in \eqref{eqn:exactmutrate} by a step function,
which will be a good approximation if the patch is large relative the scale over which $p(x)$ goes from 0 to $p_e$,
or if $p(x)$ is close to $p_e$ at some point and is symmetric about the edge of the patch. \linelabel{rr:p_estab}
We examine this more generally via exact calculation of $p(x)$ in the section \nameref{apx:establishment_sims}.

@@

Since each individual has on average $e^{-s_m}$ offspring,
the expected family size after time $t$ is $e^{- t s_m}$.
Therefore, if we define $k_e(t)$ to be the chance that the family
has gone extinct by time $t$, 
and let $K_t$ be the (random) family size conditioned on nonextinction,
then by Bayes' rule, $e^{-s_m t} = (1-k_e(t))\E[K_t]$.
It turns out that if the distribution of offspring numbers is not too heavy-tailed 
(see \citet{jagers1975branching} for details),
then $K_t$ has a limiting distribution: $K_t \dconv K$,
and 
\begin{align} \label{eqn:EK_limit}
    e^{-s_m t}/(1-k_e(t)) \to \E[K] \quad \text{as} \quad t \to \infty .
\end{align}
In other words, the chance of survival to time $t$ is a constant multiplied by $e^{-s_m t}$,
and conditional on survival, the family size is given by a fixed distribution.

@@

This suggests that 
\begin{align} \label{eqn:ratio_K}
  \migrate(x) 
    &= \rho q(x) \times 
    \frac{
        \E[1-(1-p_e)^K]
    }{
        \E[K]/2s_m
    } .
\end{align}
There are now two extremes to consider.
If $\P\{K>1/p_e\}$ is small, then each family has a small chance of establishment,
and the numerator is approximately $p_e \E[K]$.
On the other hand, if $p_e$ is not too small,
then the numerator would be close to 1, as any family that arrives will likely establish.
Combining these, and absorbing the term $\E[W|W>1]$ from equation \eqref{eqn:mean_K} 
into the constant $C$,
\begin{align} \label{eqn:twocase_migrate}
    \migrate(x) &\approx C \rho s_m \min(s_m,p_e) \left( |y| \sqrt{2 s_m}/\sigma \right)^{-(d-1)/2} \exp( - |y| \sqrt{2 s_m} / \sigma) .
%     \begin{cases}
%         q(x) \times 4 \rho s_m s_p / \xi^2  \qquad & \text{ if } s_p < s_m \\
%         q(x) \times 2 \rho s_m^2 \qquad & \text{ if } s_p > s_m 
%     \end{cases} .
\end{align}
When we use this, we set $p_e = 2 s_p / \xi^2$.

%%%%%%%%%
\subsection{What are we assuming?}
\label{ss:assumptions}

A quick review of the parameter regime we are studying here:
First, we assume that 
\textbf{(a)} each patch is large enough to support a stable population of $B$ alleles
($A^{1/d} > (\sigma/\sqrt{2s_p}) \tan^{-1}(\sqrt{s_m/s_p})$).
Next, in using equation \eqref{eqn:eqfreq}, we assume that 
\textbf{(b)} the inter-patch distance is at least several multiples of the dispersal distance
($R \gg \sigma$),
and that 
\textbf{(c)} local drift is not so strong that the equilibirum frequency is at least approximately attained
($s_m \gg 1/(\rho \sigma^d)$).
We also neglect the time for migration-selection equilibrium to be reached,
which has the effect of over-estimating $\migrate$.

We also assume that migration to a new patch takes place over a number of generations;
if there are sufficient rare, long-distance migration events
that would move between patches in a single hop,
this would require a different analysis.

@@

We will assume that once a $B$ allele is introduced (by migration or mutation) 
it becomes established in the poor habitat patch rapidly
if it escapes loss by demographic stochasticity.
Under this approximation, and the ``selective exclusion'' just discussed,
after one of the patches becomes colonized by a single $B$ allele, 
the other patch will become colonized by migration or mutation, but not by both. 
As such, the question of how the second patch adapts
simply comes down to whether a new mutation or a migrant allele is the first to become established in the second patch. 
To work out how adaptation is likely to proceed, 
we can compare expressions \eqref{eqn:mutrate} and \eqref{eqn:migrate} above
for the effective rates of adaptation by new mutation and by migration.
We work in one dimension, as the square root term appearing for two dimensions is relatively unimportant 
(as seen in figure~\ref{fig:sim_occupation_freqs} by the fact that the $d=2$ curve is nearly straight).
\linelabel{rr:sqrt_term}

@@

If the mutational influx is higher -- say, $\rho A \mu = .01$ --
the required separation between patches is only reduced to $R/\sigma > 7.3$ with $s_m=.05$.
If $s_m=.001$, then with this higher mutational influx this calculation predicts that mutation will \emph{always} be faster than migration
-- however, this should be taken very cautiously,
since as discussed above, this model does not hold if $R$ is of the same order as $\sigma$
or if $s_m$ is small enough that local drift is more important.
\linelabel{rr:separated_patches}

@@

% \begin{figure}[ht]
%   \begin{center}
%     \includegraphics{phase-diagram-log}
%   \end{center}
%   \caption{
%   Shaded contour plots of
%   mean time until adaptation in a new patch occurs by migration ($\Tmig$, blue) or by mutation ($\Tmut$, red),
%   as a function of population density ($\rho$), 
%   distance to an already adapted patch in units of dispersal distance ($R/\sigma$),
%   strength of deleterious selection ($s_m$),
%   and mutation rate scaled by patch area ($\mu A$).
%   The right-hand plot shows that $\migrate$ is relatively insensitive to $s_m$ (thanks to the square root).
%   The black, dotted line shows values where $\Tmig=\Tmut$.
%   Default parameter values were $A=100$, $\rho=10$, $\mu=10^{-8}$, $s_p=0.1$, $s_m=0.01$, $R/\sigma=20$, and $\xi^2=1$.
%   Note that population density $\rho$ and mutation influx $A\mu$ are shown on a log scale.
%   \label{fig:phase_diagram}
%   }
% \end{figure}

@@

If not yet adapted, a focal patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $R_1, \ldots, R_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}
if patches do not interfere with each other, \linelabel{rr:interference}
is
\begin{equation} \label{eqn:total_migrate}
  \sum_{i=1}^k \migrate(R_i) = C \frac{ 4 \rho s_p s_m }{\xi^2} \sum_{i=1}^{k} \left(R_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m} /\sigma\right).
\end{equation}
If the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}   \label{eqn:migprob}
  \frac{\left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(R_i \right)^{-(d-1)/2} \exp\left(- R_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
Equations \eqref{eqn:total_migrate} and \eqref{eqn:migprob} specify the transition rates of the process.
Since the compound parameter $s_p \rho / \xi^2$ is common to both migration and mutation rates,
it functions only as a time scaling of the process, 
and therefore has no effect on the final configuration, namely, 
the sets of patches sharing the genetic basis of their adaptive response.

@@

\includereviews
